---
title: "Analysis"
---

```{r}
#| message: false
#| warning: false

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(forecast)

# Raw data loading
# illerin_aldıgı_goc2: skip 3 header rows so column assignment works directly
illerin_aldıgı_goc2 <- read_excel("data/aldigi_goc.xls",  sheet = 1,
                                   skip = 3, col_names = FALSE)
illerin_verdigi_goc  <- read_excel("data/verdigi_goc.xls", sheet = 1,
                                   col_names = FALSE)
crime  <- read_excel("data/suc.xls", sheet = 1)
sucsuc <- crime
```

---

## Data Preparation

### Turkish Character Normalization

```{r}
#| message: false
#| warning: false

il_temizle <- function(x) {
  x <- trimws(x)
  x <- gsub("İ", "I", x); x <- gsub("ı", "i", x)
  x <- gsub("ğ", "g", x); x <- gsub("Ğ", "G", x)
  x <- gsub("ü", "u", x); x <- gsub("Ü", "U", x)
  x <- gsub("ş", "s", x); x <- gsub("Ş", "S", x)
  x <- gsub("ö", "o", x); x <- gsub("Ö", "O", x)
  x <- gsub("ç", "c", x); x <- gsub("Ç", "C", x)
  toupper(x)
}
```

### In-Migration (`goc_alan`)

```{r}
#| message: false
#| warning: false

goc_alan <- illerin_aldıgı_goc2

colnames(goc_alan) <- c(
  "Yil", "Il", "Toplam", "Tayin", "IsBulma", "Egitim",
  "Aile", "Konut", "AileBagimli", "MemleketeDonus",
  "Saglik", "EvAlma", "Emeklilik", "Diger", "Bilinmeyen"
)

goc_alan <- goc_alan %>%
  fill(Yil, .direction = "down") %>%
  filter(!is.na(Il),
         Il != "İl\nProvince",
         Il != "Toplam - Total",
         Il != "Toplam-Total") %>%
  mutate(
    Il_key = il_temizle(Il),
    across(c(Yil, Toplam:Bilinmeyen), as.numeric)
  )
```

### Out-Migration (`goc_veren`)

```{r}
#| message: false
#| warning: false

goc_veren <- illerin_verdigi_goc[-c(1:3), ]

colnames(goc_veren) <- c(
  "Yil", "Il", "Toplam", "Tayin", "IsBulma", "Egitim",
  "Aile", "Konut", "AileBagimli", "MemleketeDonus",
  "Saglik", "EvAlma", "Emeklilik", "Diger", "Bilinmeyen"
)

goc_veren <- goc_veren %>%
  fill(Yil, .direction = "down") %>%
  filter(!is.na(Il),
         Il != "İl\nProvince",
         Il != "Toplam - Total",
         Il != "Toplam-Total") %>%
  mutate(
    Il_key = il_temizle(Il),
    across(c(Yil, Toplam:Bilinmeyen), as.numeric)
  )
```

### Crime Data

```{r}
#| message: false
#| warning: false

crime_clean <- crime[-1, ]
colnames(crime_clean)[1] <- "Yil"

crime_long <- crime_clean %>%
  select(-2) %>%
  pivot_longer(cols = -Yil, names_to = "Il", values_to = "SucSayisi") %>%
  mutate(
    Il     = gsub("-.*", "", Il),
    Il_key = il_temizle(Il),
    Yil    = as.numeric(Yil),
    SucSayisi = as.numeric(SucSayisi)
  )

heat_df <- crime_long %>% rename(Suc = SucSayisi)
```

### Combined Analysis Table

```{r}
#| message: false
#| warning: false

crime_clean2 <- crime_long %>% filter(Yil >= 2018)

analiz_df <- goc_alan %>%
  select(Yil, Il, Il_key, AlinanGoc = Toplam) %>%
  left_join(
    goc_veren %>% select(Yil, Il_key, VerilenGoc = Toplam),
    by = c("Yil", "Il_key")
  ) %>%
  left_join(
    crime_clean2 %>% select(Yil, Il_key, SucSayisi),
    by = c("Yil", "Il_key")
  ) %>%
  select(Yil, Il, AlinanGoc, VerilenGoc, SucSayisi)

head(analiz_df)
```

---

## Net Migration

### Net Migration by Province in 2024

```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 10

net_goc_2024 <- goc_alan %>%
  filter(Yil == 2024) %>%
  select(Yil, Il, AlinanGoc = Toplam) %>%
  left_join(
    goc_veren %>% filter(Yil == 2024) %>% select(Yil, Il, VerilenGoc = Toplam),
    by = c("Yil", "Il")
  ) %>%
  mutate(NetMigration = AlinanGoc - VerilenGoc) %>%
  arrange(desc(NetMigration))

ggplot(net_goc_2024,
       aes(x = reorder(Il, NetMigration), y = NetMigration,
           fill = NetMigration > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ggtitle("Net Migration by Province in 2024") +
  xlab("Province") + ylab("Net Migration") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("red", "steelblue"), guide = "none") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7),
        plot.title  = element_text(size = 16, face = "bold", hjust = 0.5))
```

### Net Migration by Province (Last 10 Years)

```{r}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 10

son_yil <- max(goc_alan$Yil, na.rm = TRUE)
ilk_yil <- son_yil - 9

net_goc_10yil <- goc_alan %>%
  filter(Yil >= ilk_yil, Yil <= son_yil) %>%
  group_by(Il) %>%
  summarise(AlinanGoc = sum(Toplam, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    goc_veren %>%
      filter(Yil >= ilk_yil, Yil <= son_yil) %>%
      group_by(Il) %>%
      summarise(VerilenGoc = sum(Toplam, na.rm = TRUE), .groups = "drop"),
    by = "Il"
  ) %>%
  mutate(NetMigration = AlinanGoc - VerilenGoc) %>%
  arrange(desc(NetMigration))

ggplot(net_goc_10yil,
       aes(x = reorder(Il, NetMigration), y = NetMigration,
           fill = NetMigration > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ggtitle(paste0("Net Migration by Province (", ilk_yil, "–", son_yil, ")")) +
  xlab("Province") + ylab("Net Migration") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("red", "steelblue"), guide = "none") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7),
        plot.title  = element_text(size = 16, face = "bold", hjust = 0.5))
```

---

## Correlation & Regression: Migration vs. Crime

```{r}
#| message: false
#| warning: false

analiz_korelasyon <- analiz_df %>% filter(!is.na(SucSayisi))

cat("SUÇ SAYISI vs ALINAN GÖÇ\n")
print(cor.test(analiz_korelasyon$SucSayisi, analiz_korelasyon$AlinanGoc))

cat("\nSUÇ SAYISI vs VERİLEN GÖÇ\n")
print(cor.test(analiz_korelasyon$SucSayisi, analiz_korelasyon$VerilenGoc))

cat("\nREGRESSION: ALINAN GÖÇ ~ SUÇ\n")
print(summary(lm(AlinanGoc ~ SucSayisi, data = analiz_korelasyon)))

cat("\nREGRESSION: VERİLEN GÖÇ ~ SUÇ\n")
print(summary(lm(VerilenGoc ~ SucSayisi, data = analiz_korelasyon)))
```

### Scatter Plots

```{r}
#| message: false
#| warning: false

ggplot(analiz_korelasyon, aes(x = SucSayisi, y = AlinanGoc)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  ggtitle("Relationship Between Crime Count and In-Migration") +
  xlab("Crime Count") + ylab("In-Migration") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) +
  theme_minimal()

ggplot(analiz_korelasyon, aes(x = SucSayisi, y = VerilenGoc)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  ggtitle("Relationship Between Crime Count and Out-Migration") +
  xlab("Crime Count") + ylab("Out-Migration") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```

---

## Crime Trends in Major Provinces

```{r}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

suc_long <- sucsuc %>%
  rename(Yil = 1) %>%
  select(-2) %>%
  pivot_longer(cols = -Yil, names_to = "Il", values_to = "HukumluSayisi") %>%
  mutate(
    Yil = as.numeric(Yil),
    HukumluSayisi = as.numeric(HukumluSayisi)
  ) %>%
  filter(!is.na(Yil), !is.na(HukumluSayisi))

buyuk_iller <- c("İstanbul-34", "Ankara-6", "İzmir-35", "Antalya-7", "Bursa-16")

suc_trend <- suc_long %>% filter(Il %in% buyuk_iller, Yil >= 2010)

ggplot(suc_trend, aes(x = Yil, y = HukumluSayisi, color = Il)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~Il, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  labs(title = "Number of Convicts Entering Penal Institutions in Major Provinces (2010–2024)",
       x = "Year", y = "Number of Convicts") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## ARIMA Forecast: Crime Until 2030

```{r}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

iller <- c("İstanbul-34", "Ankara-6", "İzmir-35", "Çorum-19")

tum_sonuclar <- data.frame()

for (il_adi in iller) {
  il_data <- suc_long %>% filter(Il == il_adi) %>% arrange(Yil)
  son_yil <- max(il_data$Yil)
  horizon <- 2030 - son_yil

  ts_data <- ts(il_data$HukumluSayisi,
                start = min(il_data$Yil), frequency = 10)

  model  <- auto.arima(ts_data, seasonal = TRUE,
                        stepwise = FALSE, approximation = FALSE)
  tahmin <- forecast(model, h = horizon)

  actual_df <- data.frame(
    Il = il_adi, Yil = il_data$Yil, Deger = il_data$HukumluSayisi,
    Tip = "Actual", Lower80 = NA, Upper80 = NA, Lower95 = NA, Upper95 = NA
  )
  forecast_df <- data.frame(
    Il = il_adi, Yil = seq(son_yil + 1, 2030),
    Deger   = as.numeric(tahmin$mean),  Tip = "Forecast",
    Lower80 = as.numeric(tahmin$lower[, 1]),
    Upper80 = as.numeric(tahmin$upper[, 1]),
    Lower95 = as.numeric(tahmin$lower[, 2]),
    Upper95 = as.numeric(tahmin$upper[, 2])
  )
  tum_sonuclar <- bind_rows(tum_sonuclar, actual_df, forecast_df)
}

ggplot(tum_sonuclar, aes(x = Yil, y = Deger)) +
  geom_ribbon(data = tum_sonuclar %>% filter(Tip == "Forecast"),
              aes(ymin = Lower95, ymax = Upper95), fill = "lightblue", alpha = 0.35) +
  geom_ribbon(data = tum_sonuclar %>% filter(Tip == "Forecast"),
              aes(ymin = Lower80, ymax = Upper80), fill = "blue", alpha = 0.25) +
  geom_line(data  = tum_sonuclar %>% filter(Tip == "Actual"),
            color = "black", linewidth = 0.9) +
  geom_point(data = tum_sonuclar %>% filter(Tip == "Actual"),
             color = "black", size = 1.5) +
  geom_line(data  = tum_sonuclar %>% filter(Tip == "Forecast"),
            color = "blue", linewidth = 1) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red") +
  facet_wrap(~Il, scales = "free_y", ncol = 2) +
  scale_x_continuous(breaks = seq(1990, 2030, 5), limits = c(1990, 2030)) +
  scale_y_continuous(labels = comma) +
  labs(title    = "Crime Trend and Forecast Until 2030",
       subtitle = "10-year seasonal ARIMA structure tested on annual data",
       x = "Year", y = "Number of Convicts") +
  theme_minimal(base_size = 11) +
  theme(plot.title    = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text    = element_text(face = "bold"),
        axis.text.x   = element_text(angle = 45, hjust = 1))
```

---

## Ankara In-Migration Forecast

```{r}
#| message: false
#| warning: false

ankara_goc <- goc_alan %>%
  filter(Il == "Ankara") %>%
  group_by(Yil) %>%
  summarise(AlinanGoc = sum(Toplam, na.rm = TRUE), .groups = "drop") %>%
  arrange(Yil)

ankara_ts     <- ts(ankara_goc$AlinanGoc, start = min(ankara_goc$Yil), frequency = 1)
ankara_model  <- auto.arima(ankara_ts, seasonal = TRUE,
                             stepwise = FALSE, approximation = FALSE)
ankara_tahmin <- forecast(ankara_model, h = 3)

autoplot(ankara_tahmin) +
  ggtitle("Ankara In-Migration Forecast") +
  xlab("Year") + ylab("In-Migration") +
  scale_x_continuous(breaks = seq(2018, 2027, 1)) +
  theme_minimal()
```

---

## Crime Rate Heatmap

```{r}
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8

population_df <- data.frame(
  Il = c("Adana","Adiyaman","Afyonkarahisar","Agri","Aksaray","Amasya","Ankara","Antalya",
         "Ardahan","Artvin","Aydin","Balikesir","Bartin","Batman","Bayburt","Bilecik",
         "Bingol","Bitlis","Bolu","Burdur","Bursa","Canakkale","Cankiri","Corum",
         "Denizli","Diyarbakir","Duzce","Edirne","Elazig","Erzincan","Erzurum",
         "Eskisehir","Gaziantep","Giresun","Gumushane","Hakkari","Hatay","Igdir",
         "Isparta","Istanbul","Izmir","Kahramanmaras","Karabuk","Karaman","Kars",
         "Kastamonu","Kayseri","Kilis","Kirikkale","Kirklareli","Kirsehir","Kocaeli",
         "Konya","Kutahya","Malatya","Manisa","Mardin","Mersin","Mugla","Mus",
         "Nevsehir","Nigde","Ordu","Osmaniye","Rize","Sakarya","Samsun","Sanliurfa",
         "Siirt","Sinop","Sirnak","Sivas","Tekirdag","Tokat","Trabzon","Tunceli",
         "Usak","Van","Yalova","Yozgat","Zonguldak"),
  Population = c(2150000,610000,710000,540000,390000,330000,5200000,2200000,
                 100000,170000,1050000,1200000,190000,550000,80000,210000,
                 260000,340000,290000,260000,2800000,500000,180000,530000,
                 980000,1600000,350000,400000,570000,220000,760000,
                 780000,1800000,420000,140000,280000,1500000,190000,
                 420000,14000000,4000000,1050000,230000,240000,300000,
                 370000,1300000,130000,280000,350000,220000,1700000,
                 2100000,570000,760000,1350000,780000,1700000,900000,400000,
                 290000,340000,750000,500000,330000,950000,1250000,1900000,
                 310000,210000,450000,620000,900000,610000,750000,85000,
                 350000,1100000,230000,420000,600000)
)

fix_names <- function(x) {
  x <- trimws(x)
  pairs <- list(c("Ç","C"),c("Ğ","G"),c("İ","I"),c("Ö","O"),c("Ş","S"),c("Ü","U"),
                c("ç","c"),c("ğ","g"),c("ı","i"),c("ö","o"),c("ş","s"),c("ü","u"))
  for (p in pairs) x <- gsub(p[1], p[2], x)
  x
}

heat_df_rate <- heat_df %>%
  mutate(Il = fix_names(Il)) %>%
  filter(!is.na(Il), Il != "",
         !grepl("^\\.\\.\\.[0-9]+$", Il),
         !grepl("^[0-9]+$", Il)) %>%
  left_join(population_df, by = "Il") %>%
  mutate(Crime_Rate = (Suc / Population) * 100000) %>%
  filter(!is.na(Crime_Rate))

top20_rate <- heat_df_rate %>%
  group_by(Il) %>%
  summarise(Avg = mean(Crime_Rate, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Avg)) %>%
  slice_head(n = 20)

heat_top20_rate <- heat_df_rate %>% filter(Il %in% top20_rate$Il)

ggplot(heat_top20_rate,
       aes(x = factor(Yil), y = reorder(Il, Crime_Rate), fill = Crime_Rate)) +
  geom_tile(color = "white", linewidth = 0.45) +
  scale_fill_gradientn(
    colours = c("#fff7bc","#fec44f","#fd8d3c","#e31a1c","#800026"),
    trans = "log10",
    name = "Crime Rate\nper 100,000\n(log scale)"
  ) +
  labs(title    = "Crime Rate Heatmap by Province Between 2008 and 2020",
       subtitle = "Log scale — darker colors indicate higher crime rates per 100,000 population",
       x = "Year", y = "Province") +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10),
        axis.text.x   = element_text(angle = 45, hjust = 1),
        axis.text.y   = element_text(size = 10),
        panel.grid    = element_blank())
```
