---
title: "Analysis"
---

```{r setup}
#| message: false
#| warning: false
#| code-fold: false

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(forecast)
library(patchwork)
library(knitr)

# ── Dark ggplot2 theme matching the website ──────────────────────────────────
theme_dark_site <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background   = element_rect(fill = "#1e293b", color = NA),
      panel.background  = element_rect(fill = "#1e293b", color = NA),
      panel.grid.major  = element_line(color = "#2d3d52", linewidth = 0.35),
      panel.grid.minor  = element_blank(),
      axis.text         = element_text(color = "#94a3b8", size = base_size * 0.82),
      axis.title        = element_text(color = "#cbd5e1", size = base_size * 0.9),
      plot.title        = element_text(color = "#f1f5f9", face = "bold",
                                       size = base_size * 1.2, hjust = 0.5,
                                       margin = margin(b = 6)),
      plot.subtitle     = element_text(color = "#94a3b8", size = base_size * 0.88,
                                       hjust = 0.5, margin = margin(b = 10)),
      plot.caption      = element_text(color = "#64748b", size = base_size * 0.75),
      strip.background  = element_rect(fill = "#0f172a", color = NA),
      strip.text        = element_text(color = "#38bdf8", face = "bold",
                                       size = base_size * 0.88),
      legend.background = element_rect(fill = "#1e293b", color = NA),
      legend.text       = element_text(color = "#cbd5e1"),
      legend.title      = element_text(color = "#e2e8f0"),
      plot.margin       = margin(14, 14, 14, 14)
    )
}
theme_set(theme_dark_site())

COL_POS  <- "#38bdf8"   # sky-blue  — positive / in-migration
COL_NEG  <- "#f87171"   # coral-red — negative / out-migration
COL_FCST <- "#a78bfa"   # violet    — forecast
```

---

## Data Preparation

```{r load-data}
#| message: false
#| warning: false

il_temizle <- function(x) {
  x <- trimws(x)
  x <- gsub("İ","I",x); x <- gsub("ı","i",x)
  x <- gsub("ğ","g",x); x <- gsub("Ğ","G",x)
  x <- gsub("ü","u",x); x <- gsub("Ü","U",x)
  x <- gsub("ş","s",x); x <- gsub("Ş","S",x)
  x <- gsub("ö","o",x); x <- gsub("Ö","O",x)
  x <- gsub("ç","c",x); x <- gsub("Ç","C",x)
  toupper(x)
}

goc_cols <- c("Yil","Il","Toplam","Tayin","IsBulma","Egitim","Aile",
              "Konut","AileBagimli","MemleketeDonus","Saglik",
              "EvAlma","Emeklilik","Diger","Bilinmeyen")

goc_alan <- read_excel("data/aldigi_goc.xls", sheet = 1,
                        skip = 3, col_names = FALSE) %>%
  setNames(goc_cols) %>%
  fill(Yil, .direction = "down") %>%
  filter(!is.na(Il), !grepl("Province|Toplam|İl", Il, ignore.case = TRUE)) %>%
  mutate(Il_key = il_temizle(Il),
         across(c(Yil, Toplam:Bilinmeyen), ~ suppressWarnings(as.numeric(.)))) %>%
  filter(!is.na(Yil))

goc_veren <- read_excel("data/verdigi_goc.xls", sheet = 1,
                         skip = 3, col_names = FALSE) %>%
  setNames(goc_cols) %>%
  fill(Yil, .direction = "down") %>%
  filter(!is.na(Il), !grepl("Province|Toplam|İl", Il, ignore.case = TRUE)) %>%
  mutate(Il_key = il_temizle(Il),
         across(c(Yil, Toplam:Bilinmeyen), ~ suppressWarnings(as.numeric(.)))) %>%
  filter(!is.na(Yil))

crime_raw   <- read_excel("data/suc.xls", sheet = 1)
sucsuc      <- crime_raw
crime_clean <- crime_raw[-1, ]
colnames(crime_clean)[1] <- "Yil"

crime_long <- crime_clean %>%
  select(-2) %>%
  pivot_longer(cols = -Yil, names_to = "Il", values_to = "SucSayisi") %>%
  mutate(Il        = gsub("-.*", "", Il),
         Il_key    = il_temizle(Il),
         Yil       = suppressWarnings(as.numeric(Yil)),
         SucSayisi = suppressWarnings(as.numeric(SucSayisi)))

heat_df <- crime_long %>% rename(Suc = SucSayisi)

crime_2018 <- crime_long %>% filter(Yil >= 2018)

analiz_df <- goc_alan %>%
  select(Yil, Il, Il_key, AlinanGoc = Toplam) %>%
  left_join(goc_veren %>% select(Yil, Il_key, VerilenGoc = Toplam),
            by = c("Yil", "Il_key")) %>%
  left_join(crime_2018  %>% select(Yil, Il_key, SucSayisi),
            by = c("Yil", "Il_key")) %>%
  select(Yil, Il, AlinanGoc, VerilenGoc, SucSayisi)

suc_long <- sucsuc %>%
  rename(Yil = 1) %>%
  select(-2) %>%
  pivot_longer(cols = -Yil, names_to = "Il", values_to = "HukumluSayisi") %>%
  mutate(Yil = suppressWarnings(as.numeric(Yil)),
         HukumluSayisi = suppressWarnings(as.numeric(HukumluSayisi))) %>%
  filter(!is.na(Yil), !is.na(HukumluSayisi))
```

---

## Net Migration

### Net Migration by Province — 2024

```{r net-2024}
#| fig-width: 9
#| fig-height: 10

net_2024 <- goc_alan %>%
  filter(Yil == 2024) %>%
  select(Il, AlinanGoc = Toplam) %>%
  left_join(goc_veren %>% filter(Yil == 2024) %>%
              select(Il, VerilenGoc = Toplam), by = "Il") %>%
  mutate(Net = AlinanGoc - VerilenGoc) %>%
  filter(!is.na(Net))

ggplot(net_2024, aes(reorder(Il, Net), Net, fill = Net > 0)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c(COL_NEG, COL_POS), guide = "none") +
  labs(title = "Net Migration by Province — 2024",
       x = NULL, y = "Net Migration (in − out)") +
  theme(axis.text.y = element_text(size = 7))
```

### Cumulative Net Migration — Last 10 Years

```{r net-10yr}
#| fig-width: 9
#| fig-height: 10

sy <- max(goc_alan$Yil, na.rm = TRUE); iy <- sy - 9

net_10 <- goc_alan %>% filter(Yil >= iy) %>%
  group_by(Il) %>% summarise(In  = sum(Toplam, na.rm = TRUE), .groups = "drop") %>%
  left_join(goc_veren %>% filter(Yil >= iy) %>%
              group_by(Il) %>% summarise(Out = sum(Toplam, na.rm = TRUE), .groups = "drop"),
            by = "Il") %>%
  mutate(Net = In - Out) %>% filter(!is.na(Net))

ggplot(net_10, aes(reorder(Il, Net), Net, fill = Net > 0)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c(COL_NEG, COL_POS), guide = "none") +
  labs(title = paste0("Cumulative Net Migration (", iy, "–", sy, ")"),
       x = NULL, y = "Net Migration") +
  theme(axis.text.y = element_text(size = 7))
```

---

## Correlation & Regression: Crime vs. Migration

In this section, the relationship between crime count and in/out-migration across Turkish provinces is examined using Pearson correlation and simple linear regression.

**Hypotheses (In-Migration vs Crime)**

| | Hypothesis |
|---|---|
| **H₀** | There is *no* significant relationship between crime count and in-migration. |
| **H₁** | There *is* a significant relationship between crime count and in-migration. |

Since the p-value is less than 0.05, **H₀ is rejected** — a statistically significant relationship exists.

```{r corr-table}
#| message: false

df_cor <- analiz_df %>%
  filter(!is.na(SucSayisi), !is.na(AlinanGoc), !is.na(VerilenGoc))

ci  <- cor.test(df_cor$AlinanGoc,  df_cor$SucSayisi)
co  <- cor.test(df_cor$VerilenGoc, df_cor$SucSayisi)

tibble(
  Comparison     = c("In-Migration vs Crime", "Out-Migration vs Crime"),
  `Pearson r`    = round(c(ci$estimate, co$estimate), 4),
  `t-statistic`  = round(c(ci$statistic, co$statistic), 3),
  `p-value`      = format(c(ci$p.value, co$p.value), scientific = TRUE, digits = 3),
  `95% CI Lower` = round(c(ci$conf.int[1], co$conf.int[1]), 4),
  `95% CI Upper` = round(c(ci$conf.int[2], co$conf.int[2]), 4)
) %>% kable(align = c("l","r","r","r","r","r"),
            caption = "Pearson Correlation: Migration vs. Crime Count")
```

The positive correlation coefficient indicates that **provinces with higher crime counts tend to receive higher levels of migration**. However, the strength of the relationship is moderate rather than very strong — this is largely because large, economically active provinces attract more migrants *and* record higher crime counts due to higher population density and urbanisation.

### Regression Coefficients

```{r reg-table}
lm_in  <- lm(AlinanGoc  ~ SucSayisi, data = df_cor)
lm_out <- lm(VerilenGoc ~ SucSayisi, data = df_cor)

make_row <- function(m, label) {
  s <- coef(summary(m))
  r2 <- round(summary(m)$r.squared, 4)
  data.frame(Model = c(label,""), Term = rownames(s),
             Estimate = round(s[,1],3), SE = round(s[,2],3),
             `t value` = round(s[,3],3),
             `p-value` = signif(s[,4],3),
             `R²` = c(r2, NA), check.names = FALSE)
}

rbind(make_row(lm_in,"In-Migration ~ Crime"),
      make_row(lm_out,"Out-Migration ~ Crime")) %>%
  kable(align = c("l","l","r","r","r","r","r"),
        caption = "Linear Regression: Migration ~ Crime Count")
```

> **Interpretation —** The slope coefficient (~3.4) means that an increase of one unit in crime count is associated with an average increase of ~3.4 people in in-migration. The R² of ~0.13 indicates that crime count explains approximately **12.9%** of the variance in in-migration — significant but not the dominant factor. Crime should be interpreted as one of several socio-economic indicators correlated with migration, not a direct cause.

### Scatter Plots

```{r scatter}
#| fig-width: 10
#| fig-height: 4.5

p1 <- ggplot(df_cor, aes(SucSayisi, AlinanGoc)) +
  geom_point(color = COL_POS, alpha = 0.5, size = 1.4) +
  geom_smooth(method = "lm", color = "#7dd3fc", fill = "#38bdf822", linewidth = 0.9) +
  scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma) +
  labs(title = "In-Migration vs Crime", x = "Crime Count", y = "In-Migration")

p2 <- ggplot(df_cor, aes(SucSayisi, VerilenGoc)) +
  geom_point(color = COL_NEG, alpha = 0.5, size = 1.4) +
  geom_smooth(method = "lm", color = "#fca5a5", fill = "#f8717122", linewidth = 0.9) +
  scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma) +
  labs(title = "Out-Migration vs Crime", x = "Crime Count", y = "Out-Migration")

p1 + p2
```

---

## Crime Trends in Major Provinces

The following chart tracks the annual number of convicted persons entering penal institutions in five major provinces between 2010 and 2024.

```{r crime-trends}
#| fig-width: 10
#| fig-height: 6

buyuk <- c("İstanbul-34","Ankara-6","İzmir-35","Antalya-7","Bursa-16")

ggplot(suc_long %>% filter(Il %in% buyuk, Yil >= 2010),
       aes(Yil, HukumluSayisi, color = Il)) +
  geom_line(linewidth = 1.1) + geom_point(size = 1.8) +
  facet_wrap(~Il, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(2010,2024,2)) +
  scale_color_manual(values = c("#38bdf8","#a78bfa","#34d399","#fb923c","#f472b6")) +
  labs(title = "Convicted Persons Entering Penal Institutions (2010–2024)",
       x = "Year", y = "Number of Convicts") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Time Series Forecasting (ARIMA)

### Methodology

Forecasting was carried out using **ARIMA/SARIMA** models in R via the `auto.arima()` function, which automatically selects the best model order based on AIC. The general SARIMA structure is:

$$\text{ARIMA}(p,d,q)(P,D,Q)_s$$

where *p*, *d*, *q* are the non-seasonal autoregressive, differencing, and moving-average orders; *P*, *D*, *Q* are the seasonal counterparts; and *s* is the seasonal frequency. Since the dataset consists of **annual observations**, models primarily capture long-term trend behaviour. Forecasts are presented with **80% and 95% confidence intervals**.

### Forecast Plots

```{r arima-plots}
#| fig-width: 10
#| fig-height: 8

iller <- c("İstanbul-34","Ankara-6","İzmir-35","Çorum-19")
res   <- data.frame()

for (il in iller) {
  d <- suc_long %>% filter(Il == il) %>% arrange(Yil)
  if (nrow(d) < 5) next
  sy  <- max(d$Yil); h <- 2030 - sy
  ts_ <- ts(d$HukumluSayisi, start = min(d$Yil), frequency = 1)
  fit <- auto.arima(ts_, seasonal = FALSE, stepwise = TRUE, approximation = TRUE)
  fc  <- forecast(fit, h = h)
  res <- bind_rows(res,
    data.frame(Il=il, Yil=d$Yil, Val=d$HukumluSayisi, Type="Actual",
               L80=NA,U80=NA,L95=NA,U95=NA),
    data.frame(Il=il, Yil=seq(sy+1,2030),
               Val=as.numeric(fc$mean), Type="Forecast",
               L80=as.numeric(fc$lower[,1]), U80=as.numeric(fc$upper[,1]),
               L95=as.numeric(fc$lower[,2]), U95=as.numeric(fc$upper[,2])))
}

ggplot(res, aes(Yil, Val)) +
  geom_ribbon(data=filter(res,Type=="Forecast"), aes(ymin=L95,ymax=U95),
              fill=COL_FCST, alpha=0.18) +
  geom_ribbon(data=filter(res,Type=="Forecast"), aes(ymin=L80,ymax=U80),
              fill=COL_FCST, alpha=0.30) +
  geom_line(data=filter(res,Type=="Actual"),  color="#e2e8f0", linewidth=0.9) +
  geom_point(data=filter(res,Type=="Actual"), color="#e2e8f0", size=1.5) +
  geom_line(data=filter(res,Type=="Forecast"),color=COL_FCST,  linewidth=1.1) +
  geom_vline(xintercept=2020, linetype="dashed", color=COL_NEG, alpha=0.5) +
  facet_wrap(~Il, scales="free_y", ncol=2) +
  scale_x_continuous(breaks=seq(1990,2030,5), limits=c(1990,2030)) +
  scale_y_continuous(labels=comma) +
  labs(title    = "Crime Trend & ARIMA Forecast to 2030",
       subtitle = "Shaded bands: 80% (dark) and 95% (light) confidence intervals",
       x="Year", y="Number of Convicts") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

### Documented Forecast Values

The table below presents the point-estimate forecast values obtained from the ARIMA modelling exercise.

```{r forecast-table}
tibble(
  Province  = c("İstanbul","Ankara","İzmir","Çorum"),
  `2021`    = c(20923L, 13231L,  4163L, 42478L),
  `2025`    = c(23117L, 14793L,  4652L, 47325L),
  `2030`    = c(25858L, 16746L,  5262L, 53385L)
) %>%
  mutate(across(where(is.integer), ~ format(.x, big.mark=","))) %>%
  kable(align = c("l","r","r","r"),
        caption = "ARIMA Point-Estimate Forecasts — Number of Convicted Persons")
```

> **Confidence intervals (2030 estimate):** İstanbul 95% CI ≈ 16,706 – 35,011. Wider intervals reflect increasing uncertainty over longer horizons, which is expected in long-run forecasting.

### Province-Level Interpretations

| Province | Trend | Predictability |
|---|---|---|
| **İstanbul** | Strong upward; accelerating after 2010 | Moderate — wide CI |
| **Ankara** | Gradual, stable increase | High — narrow CI |
| **İzmir** | Moderate upward, flattening post-2020 | Moderate |
| **Çorum** | Sharp increase after 2014; highest relative growth | Low — very wide CI |

> All selected provinces show an **increasing long-term trend**. Because the dataset consists of annual data, the models primarily capture trend behaviour rather than short-term seasonality. Results should be treated as trend-oriented forecasts rather than exact future predictions.

---

## Crime Rate Heatmap (2008–2020)

### Objective

Rather than using raw crime counts, this analysis computes **population-adjusted crime rates** (convictions per 100,000 population) to make provinces of different sizes comparable. The top 20 provinces by average crime rate are visualised.

$$\text{Crime Rate} = \frac{\text{Convictions}}{\text{Population}} \times 100{,}000$$

```{r heatmap-data}
#| message: false

population_df <- tibble(
  Il = c("Adana","Adiyaman","Afyonkarahisar","Agri","Aksaray","Amasya",
         "Ankara","Antalya","Ardahan","Artvin","Aydin","Balikesir","Bartin",
         "Batman","Bayburt","Bilecik","Bingol","Bitlis","Bolu","Burdur",
         "Bursa","Canakkale","Cankiri","Corum","Denizli","Diyarbakir","Duzce",
         "Edirne","Elazig","Erzincan","Erzurum","Eskisehir","Gaziantep",
         "Giresun","Gumushane","Hakkari","Hatay","Igdir","Isparta","Istanbul",
         "Izmir","Kahramanmaras","Karabuk","Karaman","Kars","Kastamonu",
         "Kayseri","Kilis","Kirikkale","Kirklareli","Kirsehir","Kocaeli",
         "Konya","Kutahya","Malatya","Manisa","Mardin","Mersin","Mugla","Mus",
         "Nevsehir","Nigde","Ordu","Osmaniye","Rize","Sakarya","Samsun",
         "Sanliurfa","Siirt","Sinop","Sirnak","Sivas","Tekirdag","Tokat",
         "Trabzon","Tunceli","Usak","Van","Yalova","Yozgat","Zonguldak"),
  Pop = c(2150000,610000,710000,540000,390000,330000,5200000,2200000,
          100000,170000,1050000,1200000,190000,550000,80000,210000,
          260000,340000,290000,260000,2800000,500000,180000,530000,
          980000,1600000,350000,400000,570000,220000,760000,780000,
          1800000,420000,140000,280000,1500000,190000,420000,14000000,
          4000000,1050000,230000,240000,300000,370000,1300000,130000,
          280000,350000,220000,1700000,2100000,570000,760000,1350000,
          780000,1700000,900000,400000,290000,340000,750000,500000,
          330000,950000,1250000,1900000,310000,210000,450000,620000,
          900000,610000,750000,85000,350000,1100000,230000,420000,600000))

fix_tr <- function(x) {
  pairs <- list(c("Ç","C"),c("Ğ","G"),c("İ","I"),c("Ö","O"),c("Ş","S"),c("Ü","U"),
                c("ç","c"),c("ğ","g"),c("ı","i"),c("ö","o"),c("ş","s"),c("ü","u"))
  x <- trimws(x)
  for (p in pairs) x <- gsub(p[1],p[2],x)
  x
}

heat_rate <- heat_df %>%
  mutate(Il = fix_tr(Il)) %>%
  filter(!is.na(Il), Il != "",
         !grepl("^\\.\\.\\.[0-9]+$|^[0-9]+$", Il)) %>%
  left_join(population_df, by = "Il") %>%
  mutate(Rate = (Suc / Pop) * 1e5) %>%
  filter(!is.na(Rate))

top20 <- heat_rate %>%
  group_by(Il) %>% summarise(Avg = mean(Rate, na.rm=TRUE), .groups="drop") %>%
  slice_max(Avg, n=20)
```

### Standard Crime Rate Heatmap

```{r heatmap-std}
#| fig-width: 11
#| fig-height: 8

ggplot(heat_rate %>% filter(Il %in% top20$Il),
       aes(factor(Yil), reorder(Il, Rate), fill = Rate)) +
  geom_tile(color = "#0f172a", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = c("#0f172a","#1e3a5f","#0284c7","#38bdf8","#e0f2fe"),
    name = "Crime Rate\nper 100k") +
  labs(title    = "Crime Rate Heatmap — Top 20 Provinces (2008–2020)",
       subtitle = "Convictions per 100,000 population",
       x = "Year", y = NULL) +
  theme(axis.text.x  = element_text(angle=45, hjust=1),
        axis.text.y  = element_text(size=10),
        panel.grid   = element_blank(),
        legend.key.height = unit(1.2,"cm"))
```

### Log-Scale Heatmap

A small number of provinces with extremely high crime rates can dominate the colour scale in the standard heatmap, masking differences among the remaining provinces. The **log-transformed** version below compresses outlier values and makes moderate regional variation more visible.

```{r heatmap-log}
#| fig-width: 11
#| fig-height: 8

ggplot(heat_rate %>% filter(Il %in% top20$Il),
       aes(factor(Yil), reorder(Il, Rate), fill = Rate)) +
  geom_tile(color = "#0f172a", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = c("#0f172a","#1e3a5f","#0284c7","#38bdf8","#e0f2fe"),
    trans   = "log10",
    name = "Crime Rate\nper 100k\n(log scale)") +
  labs(title    = "Crime Rate Heatmap — Log Scale (2008–2020)",
       subtitle = "Log transformation reveals differences hidden by extreme values",
       x = "Year", y = NULL) +
  theme(axis.text.x  = element_text(angle=45, hjust=1),
        axis.text.y  = element_text(size=10),
        panel.grid   = element_blank(),
        legend.key.height = unit(1.2,"cm"))
```

### Key Findings

- Crime intensity is **not evenly distributed** — provinces such as Çorum, Amasya, Burdur, Kilis, and Artvin consistently show higher normalised crime rates.
- A **general upward trend** is visible after approximately 2013, with stronger increases between 2017 and 2020.
- **Smaller and medium-sized provinces can have higher per-capita crime rates** than major cities — highlighting the importance of population normalisation over raw counts.
- The log-scale transformation provides a more balanced view, making regional differences clearly distinguishable.

---

## Ankara In-Migration Forecast

```{r ankara-fcst}
#| fig-width: 9
#| fig-height: 5

ankara <- goc_alan %>%
  filter(Il == "Ankara") %>%
  group_by(Yil) %>% summarise(Goc = sum(Toplam, na.rm=TRUE), .groups="drop") %>%
  arrange(Yil)

if (nrow(ankara) >= 4) {
  fit_a <- auto.arima(ts(ankara$Goc, start=min(ankara$Yil), frequency=1),
                       seasonal=FALSE, stepwise=TRUE, approximation=TRUE)
  fc_a  <- forecast(fit_a, h=3)
  autoplot(fc_a) +
    scale_x_continuous(breaks=seq(2018,2027,1)) +
    scale_y_continuous(labels=comma) +
    labs(title="Ankara In-Migration Forecast (ARIMA)",
         x="Year", y="In-Migration") +
    theme_dark_site() +
    theme(axis.text.x = element_text(angle=45, hjust=1))
}
```
