---
title: "Data"
---

## Data Sources

All datasets are obtained from **TÜİK** (Türkiye İstatistik Kurumu — Turkish Statistical Institute). The analyses focus on the period **2018–2024**, which is the overlapping timeframe for all datasets.

---

## Datasets

### 1. Provincial In-Migration by Reason (`aldigi_goc.xls`)

Records the number of people who migrated **into** each province, broken down by reason for migration (work, education, family, housing, etc.).

- **Sheet:** `t10`
- **Coverage:** 81 provinces × 2018–2024
- **Key columns:** Year, Province, Total in-migration, and migration reason categories

### 2. Provincial Out-Migration by Reason (`verdigi_goc.xls`)

Records the number of people who migrated **out of** each province, broken down by reason for migration.

- **Sheet:** `t11`
- **Coverage:** 81 provinces × 2018–2024
- **Key columns:** Year, Province, Total out-migration, and migration reason categories

### 3. Education Status by Province (`egitim.xlsx`)

Population aged 6 and above classified by educational attainment level, broken down by province, year, and gender.

- **Sheet:** `Eğitim Durumu`
- **Coverage:** 81 provinces × multiple years (2018–2024)
- **Key columns:** Year, Province, Education level, Total, Male, Female, Male %, Female %

### 4. Crime Data — Convicted Persons by Province (`suc.xls`)

Number of convicted persons entering correctional institutions, broken down by province.

- **Coverage:** Multiple years including 2018 onwards
- **Key columns:** Province (columns), Year (rows), Count of convicted persons

---

## Data Preview

```{r}
#| message: false
#| warning: false

library(readxl)
library(knitr)

aldigi_goc  <- read_excel("data/aldigi_goc.xls",  sheet = "t10")
verdigi_goc <- read_excel("data/verdigi_goc.xls", sheet = "t11")
egitim      <- read_excel("data/egitim.xlsx",      sheet = "Eğitim Durumu")
suc         <- read_excel("data/suc.xls")
```

### In-Migration Data

```{r}
kable(head(aldigi_goc, 10))
```

### Out-Migration Data

```{r}
kable(head(verdigi_goc, 10))
```

### Education Status Data

```{r}
kable(head(egitim, 10))
```

### Crime Data

```{r}
kable(head(suc, 10))
```
