---
title: "Data"
---

## Data Sources

All datasets are obtained from **TÜİK** (Türkiye İstatistik Kurumu — Turkish Statistical Institute). The analyses focus on the period **2018–2024**, which is the overlapping timeframe for all datasets.

---

## Datasets

### 1. Provincial In-Migration by Reason (`aldigi_goc.xls`)

Records the number of people who migrated **into** each province, broken down by reason for migration.

- **Coverage:** 81 provinces × 2018–2024
- **Key columns:** Year, Province, Total, and 12 migration-reason categories

### 2. Provincial Out-Migration by Reason (`verdigi_goc.xls`)

Records the number of people who migrated **out of** each province, broken down by reason for migration.

- **Coverage:** 81 provinces × 2018–2024
- **Key columns:** Year, Province, Total, and 12 migration-reason categories

### 3. Education Status by Province (`egitim.xlsx`)

Population aged 6 and above classified by educational attainment level, broken down by province, year, and gender.

- **Coverage:** 81 provinces × 2018–2024
- **Key columns:** Year, Province, Education level, Total, Male, Female

### 4. Crime Data — Convicted Persons by Province (`suc.xls`)

Number of convicted persons entering correctional institutions, broken down by province and year.

- **Coverage:** 81 provinces × 2008–2020+
- **Key columns:** Year (rows) × Province (columns)

---

## Data Preview

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(readxl)
library(dplyr)
library(tidyr)
library(knitr)

# shared column names for both migration files
goc_cols <- c("Year", "Province", "Total", "Appointment", "FindingWork",
              "Education", "Family", "Housing", "FamilyDependent",
              "ReturnHometown", "Health", "HousePurchase", "Retirement",
              "Other", "Unknown")

clean_preview <- function(df, n = 8) {
  df %>%
    fill(Year, .direction = "down") %>%
    filter(!is.na(Province),
           !grepl("Province|Total|İl|Toplam", Province, ignore.case = TRUE)) %>%
    mutate(across(everything(), function(x) suppressWarnings(as.numeric(x)),
                  .names = "{.col}"),
           .keep = "none") %>%
    { df %>%
        fill(Year, .direction = "down") %>%
        filter(!is.na(Province),
               !grepl("Province|Total|İl|Toplam", Province, ignore.case = TRUE)) %>%
        mutate(Year = suppressWarnings(as.numeric(Year))) %>%
        filter(!is.na(Year)) %>%
        head(n)
    }
}
```

### In-Migration Data (first 8 rows)

```{r}
#| message: false
#| warning: false

goc_in_raw <- read_excel("data/aldigi_goc.xls",  sheet = 1,
                          skip = 3, col_names = FALSE)
colnames(goc_in_raw) <- goc_cols

goc_in_raw %>%
  fill(Year, .direction = "down") %>%
  filter(!is.na(Province),
         !grepl("Province|Total|İl|Toplam", Province, ignore.case = TRUE)) %>%
  mutate(Year = suppressWarnings(as.numeric(Year))) %>%
  filter(!is.na(Year)) %>%
  select(Year, Province, Total, FindingWork, Education, Family, Housing) %>%
  mutate(across(where(is.numeric), ~ format(.x, big.mark = ",", scientific = FALSE))) %>%
  head(8) %>%
  kable(col.names = c("Year", "Province", "Total", "Work", "Education",
                      "Family", "Housing"),
        align = c("c","l","r","r","r","r","r"),
        caption = "Provincial In-Migration by Selected Reason (aldigi_goc.xls)")
```

### Out-Migration Data (first 8 rows)

```{r}
#| message: false
#| warning: false

goc_out_raw <- read_excel("data/verdigi_goc.xls", sheet = 1,
                           skip = 3, col_names = FALSE)
colnames(goc_out_raw) <- goc_cols

goc_out_raw %>%
  fill(Year, .direction = "down") %>%
  filter(!is.na(Province),
         !grepl("Province|Total|İl|Toplam", Province, ignore.case = TRUE)) %>%
  mutate(Year = suppressWarnings(as.numeric(Year))) %>%
  filter(!is.na(Year)) %>%
  select(Year, Province, Total, FindingWork, Education, Family, Housing) %>%
  mutate(across(where(is.numeric), ~ format(.x, big.mark = ",", scientific = FALSE))) %>%
  head(8) %>%
  kable(col.names = c("Year", "Province", "Total", "Work", "Education",
                      "Family", "Housing"),
        align = c("c","l","r","r","r","r","r"),
        caption = "Provincial Out-Migration by Selected Reason (verdigi_goc.xls)")
```

### Education Status Data (first 8 rows)

```{r}
#| message: false
#| warning: false

egitim_raw <- read_excel("data/egitim.xlsx", sheet = 1, col_names = FALSE)

# Find first row where column 1 looks like a 4-digit year
year_row <- which(sapply(egitim_raw[[1]],
                         function(x) grepl("^20[0-9]{2}$", as.character(x))))[1]
skip_n   <- if (!is.na(year_row) && year_row > 1) year_row - 1L else 2L

egitim <- read_excel("data/egitim.xlsx", sheet = 1, skip = skip_n) %>%
  filter(!is.na(.[[1]]))

egitim %>%
  head(8) %>%
  kable(caption = "Education Status by Province (egitim.xlsx)")
```

### Crime Data — Sample (5 provinces, recent years)

```{r}
#| message: false
#| warning: false

suc_raw <- read_excel("data/suc.xls", sheet = 1)

# Row 1 is typically a second header row; keep from row 2 onwards
suc_clean <- suc_raw[-1, ]
colnames(suc_clean)[1] <- "Year"

# Remove the 2nd column if it contains duplicate info
if (ncol(suc_clean) > 1 && all(is.na(suc_clean[[2]]) | suc_clean[[2]] == "")) {
  suc_clean <- suc_clean[, -2]
}

suc_preview <- suc_clean %>%
  mutate(Year = suppressWarnings(as.numeric(Year))) %>%
  filter(!is.na(Year)) %>%
  arrange(desc(Year)) %>%
  head(6) %>%
  select(1:6)   # Year + first 5 provinces

kable(suc_preview,
      caption = "Convicted Persons by Province — Sample (suc.xls)",
      align = c("c", rep("r", ncol(suc_preview) - 1)))
```
